---
title: "Hello JS in R Markdown"
author: "Alicia Schep"
date: "2018-02-10"
slug: js-rmarkdown
categories: ["R"]
tags: ["R Markdown", "javascript", "visualization", "knitr"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = TRUE)
```

One of many rad features of R Markdown is that, despite the name, you can add chunks in languages other than R, including python, bash, Rcpp, and SQL. Until recently, I hadn't realized that you could also add javascript chunks &mdash; on my version of Rstudio, it doesn't come up as an option on the insert chunk dropdown. Simply replacing the `r` at the start of the chunk with `js` will insert the code within the chunk into script tags in the document. As I'm in the process of learning d3 and exploring various javascript visualization libraries, I was curious to try this feature out.     

A [blog post from Nick Strayer](http://livefreeordichotomize.com/2017/01/24/custom-javascript-visualizations-in-rmarkdown/) was really helpful for getting started with JS in R Markdown. The main change I wanted to make from the approach in that post was to place the plot below the code, similar to what happens with R code. After figuring out a quick way to do this, I ended up becoming interested in how knitr's language engines work, and was pleasantly surprised by how accessible the engines are &mdash; with a few lines of code you can add a new chunk option to affect the output of a javascript chunk!

In this post I'll go over the basic of adding javascript code to R Markdown and conclude with a modified javascript knitr engine.

## Sending data from R to javascript

If you want to be able to access data from R in a javascript chunk, you'll need to explicitly pass that data to javascript. Nick's blog post includes a handy little function for passing a data frame from R to javascript by converting it to JSON and then wrapping it in a `<script>` tag. I modified this function to enable specifying the variable name and to take advantage of the htmltools tags functionality so that the chunk does not have to be set with option `results='asis'`:

```{r preliminary_send_df}
# Function adapted from Nick Strayer's at 
# http://livefreeordichotomize.com/2017/01/24/custom-javascript-visualizations-in-rmarkdown/
df_to_js <- function(x, var_name = "data", ...){
  
  json_data <- jsonlite::toJSON(x, ...)
  
  htmltools::tags$script(paste0("var ",var_name," = ", json_data, ";"))
}

df_to_js(iris)
```

## Importing the javascript libraries.  

To import javascript libraries, you can add script tags to the body of the R markdown in plain html. Here I wanted to make a plot using dimple.js (which also requires d3). To add it in via an R code chunk so that it shows up in a chunk, the htmltools package is again useful:

```{r sending, results='asis'}
htmltools::tagList(
  htmltools::tags$script(src = "https://d3js.org/d3.v4.min.js"),
  htmltools::tags$script(src = "https://cdnjs.cloudflare.com/ajax/libs/dimple/2.3.0/dimple.latest.min.js")
)
```

## Javascript chunk

A javascript chunk is made in R Markdown by specifying "js" instead of "r" as the language in the chunk header. It can also be useful to set the "class.source" chunk option in the header:

    ```{js, class.source = "jsvis1"}`r ''`
    // javascript code goes here
    ```

The "class.source" option will set the class of the the source code chunk.  Assigning that chunk a class name makes it easier to insert a 'div' from within the javascript to hold your visualization *after* the source chunk[^1] so that it appears in the right place in the document (or at least behaves similarly to a plot created by an R chunk).

We'll use jquery to insert the div:

```{r}
htmltools::tags$script(src = "https://cdnjs.cloudflare.com/ajax/libs/jquery/3.3.1/jquery.min.js")
```

The javascript:

```{js, class.source = "jsvis1"}
// Create the div that the figure will be made in 
$('<div id="vis"></div>').insertAfter(".jsvis1");

// Make the visualization
var svg = dimple.newSvg("#vis", "100%", 600);
var chart = new dimple.chart(svg, data);
chart.addMeasureAxis("x", "Sepal.Length");
chart.addMeasureAxis("y", "Petal.Width");
chart.addSeries(["Sepal.Length","Petal.Width","Species"], dimple.plot.bubble);
chart.draw();

// Responsive sizing
window.addEventListener('resize', function(event){
  chart.draw(0, true);
});
```

Woohoo! The figure showed up below the code.

## Modifiying the knitr language engine

I was curious if there was a way to add that extra div by specifying a custom option to the knitr chunk. That ended up being easier than I was expecting thanks to the incredible control knitr lets you take over how chunks in different languages are executed.  

I took a look at what the default knitr "engine" for javascript was by looking up `knitr::knit_engines$get("js")`. I edited that to take into consideration an extra chunk option, div. If that argument is present, a div is created in front of the output with that id. 

```{r}
knitr::knit_engines$set("js" = function(options){
    out = if (options$eval) {
        div_pre <- if (!is.null(options$div)) 
          paste0('<div id="',options$div,'"></div>')
        code = c(div_pre,
                '<script type="text/javascript">',
                 options$code,
                 '</script>')
        paste(code, collapse = "\n")
    }
    options$results = "asis"
    knitr::engine_output(options, options$code, out)
})
```

For the next chunk we set it up like:

    ```{js, div = "vis2"}`r ''`
    // javascript code goes here
    ```

With the code:

```{js, div = "vis2"}
// Make the visualization
var svg2 = dimple.newSvg("#vis2", "100%", 600);
var chart2 = new dimple.chart(svg2, data);
chart2.addMeasureAxis("x", "Sepal.Width");
chart2.addMeasureAxis("y", "Petal.Length");
chart2.addSeries(["Sepal.Width","Petal.Length","Species"], dimple.plot.bubble);
chart2.draw();

// Responsive sizing
window.addEventListener('resize', function(event){
  chart2.draw(0, true);
});
```


## The downsides of JS in R Markdown

The approach described above for incorporating Javascript in R Markdown works pretty well for incorporating a snippet of javascript into an R Markdown document. However, when editing the document interactively in RStudio, you won't be able to see the immediate output the way you would for R plots &mdash; you can only see the output after knitting the document. 

[^1]: Just moving the 'div' element below the chunk doesn't work, because it won't exist when the code is first run

